---
title: 'Visualizing Locke and Mill: A Tidy Text Analysis'
author: "Seth Dobson"
date: "5/17/2020"
output:
  md_document:
    variant: markdown_github
---

## Introduction

In this post, I will perform a text mining analysis of two important philosophical works on the topic of individual liberty: John Locke's [*Second Treatise of Governement*](https://en.wikipedia.org/wiki/Two_Treatises_of_Government) and John Stuart Mill's [*On Liberty*](https://en.wikipedia.org/wiki/On_Liberty). 

I will rely mainly on the [tidytext](https://cran.r-project.org/web/packages/tidytext/vignettes/tidytext.html) package, as described in the book [*Text Mining with R: A Tidy Approach*](https://www.tidytextmining.com/) written by [Julia Silge](https://juliasilge.com/) and [David Robinson](http://varianceexplained.org/about/). My focus will be on comparing word frequencies to understand the degree to which word usage correlates between Locke and Mill.

## A brief excursion into political philosophy

*Second Treatise of Government* was published in 1690, while *On Liberty* was published 1859. Both of these highly influential works are ostensibly about individual freedom. However, they are nonetheless different in important ways. 

First, while *Second Treatise* is predominantly about civil liberty, or freedom from tyrannical forms of government that infringe upon individual rights to life, liberty, and property, *On Liberty* mainly focuses on personal liberty, i.e., freedom of speech/thought and social non-conformity.

Second, the two authors differed fundamentally in their view of [natural rights](https://en.wikipedia.org/wiki/Natural_and_legal_rights). The background research for this section is based on Modules 2 and 8 from the Cato Institute's [Home Study Course](https://www.cato.org/cato-university/home-study-course)"). Locke argued that natural law gives all people universal and unalienable rights of self-ownership and self-determination, and that the purpose of government is to protect these rights. Mill on the other hand, being a disciple of [Jeremy Bentham](https://en.wikipedia.org/wiki/Jeremy_Bentham), did not believe in the logical existence of natural rights, and therefore made his defense of individual liberty from the standpoint of [utilitarianism](https://en.wikipedia.org/wiki/Utilitarianism), or the greatest happiness for the greatest number.

Based on these differences, you might expect word usage to differ somewhat between *Second Treatise* and *On Liberty*. Let's use R to take a look!

## Tidytext in action

[Project Gutenberg](https://www.gutenberg.org/) has both *Second Treatise of Government* and *On Liberty* in its free digital archives. I used the [gutenbergr](https://cran.r-project.org/web/packages/gutenbergr/vignettes/intro.html) package developed by David Robinson to download the works.

```{r, message=FALSE, warning=FALSE}
# Load packages and set plot theme
library(gutenbergr)
library(tidyverse)
library(tidytext)
# library(ggthemr)
# ggthemr("dust")

```

First, I used the metadata file from the gutenbergr package to find the ID numbers for *Second Treatise* and *On Liberty*. Once I had these, I was able to download each work using the `gutenberg_download()` function.

```{r, include=FALSE}
# Get book ID numbers
gutenberg_metadata %>% 
  filter(title %in% c("Second Treatise of Government",
                      "On Liberty")) %>% 
  select(gutenberg_id, title)

```

Let's start with Locke.

```{r, message=FALSE}
# Download Second Treatise of Governement
locke <- gutenberg_download(gutenberg_id = 7370)

# Show a selection of text
locke %>% 
  select(text) %>% 
  slice(50:56)

```

After we tidy it up with the tidytext package, it looks like this.

```{r, message=FALSE}
# Tokenize text and remove stop words
tidy_locke <- locke %>% 
  unnest_tokens(output = word, input = text, token = "words") %>%
  anti_join(stop_words) %>% 
  mutate(word = str_extract(word, "[a-z']+"))

# Show a selection of tokens (words)
tidy_locke %>% 
  select(word) %>% 
  slice(144:170)

```